Effects of COVID-19 related social media use on well-being

### 2023-08-08 manuscript.tex
- shortened ms

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus?
In what region is it spreading?
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high (Statista, 2021).
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed ``doomscrolling'' emerged (Sharma et al., 2022).
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children (Klein, 2021).
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health (Sandstrom et al., 2021).
Although research has begun addressing this question
(e.g., Bendau et al., 2021; Eden et al., 2020; Sewall et al., 2021),
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are only short-term, with well-being after some time returning to prior levels (Sheldon \& Lucas, 2014).
Only very specific factors such as unemployment, disability, or death can cause long-term changes in well-being (Lucas, 2007).
So although well-being can be affected by external events and factors, this does not happen easily.

@@

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory (Zillmann, 1988), using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs (Zillmann, 1988).
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood (Marciano et al., 2022).
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.
On the other hand, although mood management theory suggests that effects should not be overly negative, it could be that although short-term effects are positive, long-term effects are negative.
Precisely because social media have so many positive consequences in the short run, this might cause problem in the long run---for example, because social media use displaces other potentially more meaningful and/or relaxing activities (Hall \& Liu, 2022).
As with many other things, there can be too much of a good thing.
It is therefore often asked whether social media can become addictive, and users sometimes express this fear themselves (Yang et al., 2021).

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
Despite incorrect information, social media provide a vast platform for disseminating \emph{accurate and timely information} about COVID-19 (John Hopkins University, 2023).
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
Social media can be a valuable \emph{educational tool}, with various organizations and experts sharing informative content about COVID-19 (World Health Organization, 2023).
Access to educational resources on social media can empower individuals to make informed decisions regarding their health.
Social media platforms enable individuals to \emph{connect with others} who are experiencing similar challenges during the pandemic (Guazzini et al., 2022).
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for \emph{maintaining mental well-being} during the pandemic (Twitter, 2020).
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
Social media campaigns and initiatives can promote \emph{positive COVID-19 behaviors}, such as mask-wearing, physical distancing, hand hygiene, and vaccination (Athey et al., 2023; Hunt et al., 2022).
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread \emph{false or misleading information} about COVID-19 (Li et al., 2020).
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
Constant exposure to COVID-19-related content on social media can lead to \emph{information overload} and contribute to heightened anxiety levels (J. Fan \& Smith, 2021).
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic (Sharma et al., 2022).
Social media algorithms are designed to show users content that aligns with their interests and beliefs, potentially contributing to \emph{echo chambers} (Cinelli et al., 2021).
In the context of COVID-19, this might reinforce false or misleading information and prevent individuals from considering alternative perspectives.
Social media platforms are known for fostering negativity, with users sometimes engaging in \emph{cyberbullying and harassment} (Giumetti \& Kowalski, 2022).
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts.
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
Social media often showcase the highlights and accomplishments of others, encouraging \emph{social comparison} (Przybylski et al., 2013).
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances (Sharma et al., 2022).

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison (Klein, 2021; Stainback et al., 2020; Yue et al., 2022).
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms (Bendau et al., 2021).
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes.
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being (Guazzini et al., 2022).
Additionally, engaging more on social media was associated with reduced feelings of loneliness (Latikka et al., 2022).
A couple of studies reported mostly neutral effects of social media use on well-being indicators (Bradley \& Howard, 2021; Eden et al., 2020; Sewall et al., 2021).
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

When using longitudinal designs to analyze causality, it is important to (a) focus on within-person effects (Hamaker, 2014); to (b) control for confounders (Rohrer \& Murayama, 2021); and to (c) test a plausible interval between measures (Dormann \& Griffin, 2015).
First, in non-experimental designs it makes much sense to analyze causal effects from an internal, within-person perspective (Hamaker, 2014).
If a specific person changes their media diet, we need to measure how this behavior affects their well-being.
Between-person comparisons from longitudinal data cannot provide such insights (Lucas, 2022).
To test the hypothesis, I thus consider only the within-person effects.

Second, to identify confounders we should control for variables that affect both media use and well-being, which helps isolate the actual effect (Rohrer, 2018).
Because we are adopting a within-person perspective, we need to implement \emph{time-varying} confounders (Rohrer \& Murayama, 2021).
And because we are determining the \emph{overall} causal effect, we need to make sure \emph{not} to control for mediating variables (Rohrer, 2018), for doing so would bias our assessment of the causal effect.
In this study, I hence preregistered to control for the following variables, which either have already been shown or are likely to affect both social media use and well-being, and which also are not mediators:
gender, age, education, Austria country of birth, Austria country of birth of parents, residency Vienna, text-based news consumption, video-based news consumption, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, disposition to take risks, and locus of control (Eger \& Maridal, 2015; Ward et al., 2016).

@@

Life satisfaction was measured with the item ``All things considered, how satisfied are you with your life as a whole nowadays?'', which comes from the European Social Survey (European Social Survey, 2021).
The response options ranged from 0 (\emph{extremely dissatisfied}) to 10 (\emph{extremely satisfied}).

@@

These specific observations notwithstanding, several general trends can be observed.
First, overall the results do suggest that effects of COVID-19 related social media use on well-being tend to take place in the negative as opposed to the positive spectrum.
Although very small, five statistically significant negative results of COVID-19 related social media use on well-being were found.
Only one positive effect emerged.
Also note that in the analyses several control variables were included, ruling out plausible alternative explanations for the negative results.
For example, it was controlled for as to whether or not participants contracted a COVID-19 infection during a specific wave.
Hence, we can rule out the alternative explanation that having an infection was the root cause of increased communication and reduced well-being.

Second, six significant outcomes emerged for positive or negative affect, but only one for life satisfaction.
Life satisfaction is more stable and not that easily affected by any type or channel of social media communication.
The more fluctuating positive and negative affect, however, were affected (albeit only slightly).
Liking, sharing, and posting COVID-19 related content, and spending more time on Twitter and YouTube to browse COVID-19 related content, all slightly negatively influenced affect.
This is aligned with prior findings which showed that social media use can trigger negative affect, but that it is less likely to determine life satisfaction (Huang, 2017).
Conversations about COVID-19 on social media are often extreme, negative, or aggressive (L. Fan et al., 2020).
More deeply engaging with this type of content could negatively affect active authors.
The hypothesis that tonality could explain the negative effects is especially supported by the observation that spending more time on Twitter and YouTube than usual increased negative affect.
Communication on both channels is often found to be negative and impolite (e.g., Mueller \& Saeltzer, 2022), also when compared to other SNSs (Halpern \& Gibbs, 2013).
Consuming more negative and misleading information could hence explain the (slightly) increased levels of negative affect.

Third, the results show that it makes sense to analyze different communication types and communication channels.
Reading slightly reduced positive affect, while liking, sharing, and posting slightly increased negative affect.
Interestingly, posting COVID-19 related comment slightly increased negative affect, while at the same time it also slightly increased positive affect.
Posting content is often met with strong reaction, both positive by means of likes and negative by means of critical comments.
Overall, though, posting led to slightly reduced levels of life satisfaction.
In conclusion, whereas it was often stated that passive use is bad and active use good (Verduyn et al., 2015), this pattern was only partially found here.
The results are aligned with the findings from Valkenburg et al. (2022), who could not confirm that active use is good and that passive use is bad.
Focusing on communication channels, Twitter and YouTube seem to be more negative, as has often been observed (Halpern \& Gibbs, 2013), while Instagram, WhatsApp, and Facebook were neutral.
But, again, all of these effects are very small.
Future research might elaborate on these specific relations to probe their stability and relevance.

Taken together, the results are hence aligned with the underlying theoretical models and prior empirical results.
The findings support the differential susceptibility of media effects model (Valkenburg \& Peter, 2013), such that effects are generally small and that they depend on the type and channel of communication.
Additional analyses did not reveal that effects depended on gender.
Age also large did not play a significant moderation role, but effects of posting COVID-19 related content were found to be more negative for Generation Z.
Indeed, it has often been argued that effects of social media use are more negative for Gen Z than for prior generations, and this finding can be seen a further tentative support for this hypothesis.
From a broader perspective, the results are well-aligned with mood management theory (Zillmann, 1988) and the uses and gratifications approach (Katz et al., 1973), whose premises preclude particularly negative effects of routine and widespread media consumption.
Both theories posit that if the effects of social media were indeed profoundly negative on average, then people likely would not spend so much time on social media engaging with COVID-19 content.
Finally, recent empirical studies and meta-analyses reported rather small negative effects, too.
Several studies found that the effects of various types of social media use on well-being are small, often too small to matter (Bendau et al., 2021; Ferguson et al., 2021; Meier \& Reinecke, 2020; Orben, 2020), echoing the results obtained here.

@@

Focusing on within-person effects and controlling for several potential confounders, this study provides an improved perspective on assessing causality.
However, several challenges remain.
In order to correctly establish causality in non-experimental designs, it is necessary to control for \emph{all} relevant confounding third variables (Rohrer, 2018).
Although this study included are large list of confounders, it could still be that crucial variables were missed.
More thought needs to be invested in which factors to control for and, equally important, for which factors not to control for.
I hope this study provides a first step into this direction.

@@

Both media use and well-being were measured using self-reports.
Because assessing well-being necessarily requires introspection, using self-reports for affect and life satisfaction is adequate.
However, for social media use objective measures are clearly preferable, as people often cannot reliably estimate their use (Scharkow, 2016).
At the same time, most objective measures cannot capture the content or the motivation of use.
Hence, for the type of research question analyzed here, it still seems necessary to use self-reported measures.
In many cases they can still be informative (Verbeij et al., 2021).

Being collected in a single country, the generalizability of the results is limited.
The results apply primarily to the more Western sphere.
They might not hold true in other cultures, especially cultures with a different media landscape or alternative social media channels.

@@

In this study, COVID-19 related social media use did not meaningfully affect well-being.
Very small negative effects were found for writing COVID-19 related posts, sharing COVID-19 related content, and spending more time than usual on Twitter.
Factors other than social media use, however, were meaningfully related to well-being, including physical health, exercise, satisfaction with democracy, or believing that one is in control of one's life.
In light of the overall very small effects, engaging in COVID 19-related social media use should not be considered a major concern for one's well-being.
Hence, when trying to improve well-being during a pandemic, instead of focusing on social media it seems more fruitful to address other, more pertinent societal problems related to health care, regular exercise, or psychological resilience.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bradleyStressMoodSmartphone2021}{}}%
Bradley, A., \& Howard, A. (2021). \emph{Stress, mood, and smartphone use in {University} students: {A} 12-week longitudinal study}. \url{https://doi.org/10.31219/osf.io/frvpb}

\leavevmode\vadjust pre{\hypertarget{ref-cinelliEchoChamberEffect2021}{}}%
Cinelli, M., De Francisci Morales, G., Galeazzi, A., Quattrociocchi, W., \& Starnini, M. (2021). The echo chamber effect on social media. \emph{Proceedings of the National Academy of Sciences}, \emph{118}(9), e2023301118. \url{https://doi.org/10.1073/pnas.2023301118}

@@

\leavevmode\vadjust pre{\hypertarget{ref-europeansocialsurveyESS9Edition20182021}{}}%
European Social Survey. (2021). \emph{{ESS9} edition 3.1 - 2018 {Documentation Report}}. \url{https://www.europeansocialsurvey.org/docs/round9/survey/ESS9_data_documentation_report_e03_1.pdf}

\leavevmode\vadjust pre{\hypertarget{ref-fanInformationOverloadWellbeing2021}{}}%
Fan, J., \& Smith, A. P. (2021). Information overload, wellbeing and {COVID-19}: {A} survey in {China}. \emph{Behavioral Sciences}, \emph{11}(5, 5), 62. \url{https://doi.org/10.3390/bs11050062}

\leavevmode\vadjust pre{\hypertarget{ref-fanStigmatizationSocialMedia2020}{}}%
Fan, L., Yu, H., \& Yin, Z. (2020). Stigmatization in social media: {Documenting} and analyzing hate speech for {\textsc{COVID}} ‐19 on {Twitter}. \emph{Proceedings of the Association for Information Science and Technology}, \emph{57}(1). \url{https://doi.org/10.1002/pra2.313}

\leavevmode\vadjust pre{\hypertarget{ref-fergusonThisMetaanalysisScreen2021}{}}%
Ferguson, C. J., Kaye, L. K., Branley-Bell, D., Markey, P., Ivory, J. D., Klisanin, D., Elson, M., Smyth, M., Hogg, J. L., McDonnell, D., Nichols, D., Siddiqui, S., Gregerson, M., \& Wilson, J. (2021). Like this meta-analysis: {Screen} media and mental health. \emph{Professional Psychology: Research and Practice}. \url{https://doi.org/10.1037/pro0000426}

@@

\leavevmode\vadjust pre{\hypertarget{ref-giumettiCyberbullyingSocialMedia2022}{}}%
Giumetti, G. W., \& Kowalski, R. M. (2022). Cyberbullying via social media and well-being. \emph{Current Opinion in Psychology}, \emph{45}, 101314. \url{https://doi.org/10.1016/j.copsyc.2022.101314}

@@

\leavevmode\vadjust pre{\hypertarget{ref-hallSocialMediaUse2022}{}}%
Hall, J. A., \& Liu, D. (2022). Social media use, social displacement, and well-being. \emph{Current Opinion in Psychology}, \emph{46}, 101339. \url{https://doi.org/10.1016/j.copsyc.2022.101339}

\leavevmode\vadjust pre{\hypertarget{ref-halpernSocialMediaCatalyst2013}{}}%
Halpern, D., \& Gibbs, J. (2013). Social media as a catalyst for online deliberation? {Exploring} the affordances of {Facebook} and {YouTube} for political expression. \emph{Computers in Human Behavior}, \emph{29}(3), 1159--1168. \url{https://doi.org/10.1016/j.chb.2012.10.008}

@@

\leavevmode\vadjust pre{\hypertarget{ref-huangTimeSpentSocial2017}{}}%
Huang, C. (2017). Time spent on social network sites and psychological well-being: {A} meta-analysis. \emph{Cyberpsychology, Behavior and Social Networking}, \emph{20}(6), 346--354. \url{https://doi.org/10.1089/cyber.2016.0758}

\leavevmode\vadjust pre{\hypertarget{ref-huntSocialMediaBased2022}{}}%
Hunt, I. de V., Dunn, T., Mahoney, M., Chen, M., Nava, V., \& Linos, E. (2022). A social media‒based public health campaign encouraging {COVID-19} vaccination across the {United States}. \emph{American Journal of Public Health}, \emph{112}(9), 1253--1256. \url{https://doi.org/10.2105/AJPH.2022.306934}

\leavevmode\vadjust pre{\hypertarget{ref-johnhopkinsuniversityCOVID19Map2023}{}}%
John Hopkins University. (2023). \emph{{COVID-19 Map}}. {Johns Hopkins Coronavirus Resource Center}. \url{https://coronavirus.jhu.edu/map.html}

@@

\leavevmode\vadjust pre{\hypertarget{ref-kittelAustrianCoronaPanel2020}{}}%
Kittel, B., Kritzinger, S., Boomgaarden, H., Prainsack, B., Eberl, J.-M., Kalleitner, F., Lebernegg, N. S., Partheymüller, J., Plescia, C., Schiestl, D. W., \& Schlogl, L. (2020). \emph{Austrian {Corona Panel Project} ({SUF} edition)} {[}Data set{]}. {AUSSDA}. \url{https://doi.org/10.11587/28KQNS}

\leavevmode\vadjust pre{\hypertarget{ref-kleinDarklySoothingCompulsion2021}{}}%
Klein, J. (2021, March 3). \emph{The darkly soothing compulsion of 'doomscrolling'}. \url{https://www.bbc.com/worklife/article/20210226-the-darkly-soothing-compulsion-of-doomscrolling}

@@

\leavevmode\vadjust pre{\hypertarget{ref-lucasWhyCrossLaggedPanel2022}{}}%
Lucas, R. E. (2022). \emph{Why the {Cross-Lagged Panel Model} is {Almost Never} the {Right Choice}} {[}Preprint{]}. {PsyArXiv}. \url{https://doi.org/10.31234/osf.io/pkec7}

@@

\leavevmode\vadjust pre{\hypertarget{ref-normanInterpretationChangesHealthrelated2003}{}}%
Norman, G., Sloan, J., \& Wyrwich, K. (2003). Interpretation of changes in health-related quality of life: {The} remarkable universality of half a standard deviation. \emph{Medical Care}, \emph{41}(5), 582--592. \href{Retrieved\%20from\%20http://www.jstor.org/stable/3768017}{Retrieved from http://www.jstor.org/stable/3768017}

\leavevmode\vadjust pre{\hypertarget{ref-orbenTeenagersScreensSocial2020}{}}%
Orben, A. (2020). Teenagers, screens and social media: A narrative review of reviews and key studies. \emph{Social Psychiatry and Psychiatric Epidemiology}, \emph{55}(4), 407--414. \url{https://doi.org/10.1007/s00127-019-01825-4}

@@

\leavevmode\vadjust pre{\hypertarget{ref-rohrerTheseAreNot2021}{}}%
Rohrer, J. M., \& Murayama, K. (2021, October 11). \emph{These are not the effects you are looking for: {Causality} and the within-/between-person distinction in longitudinal data analysis}. \url{https://doi.org/10.31234/osf.io/tg4vj}

\leavevmode\vadjust pre{\hypertarget{ref-sandstromDoomscrollingCOVIDNews2021}{}}%
Sandstrom, G., Buchanan, K., Aknin, L., \& Lotun, S. (2021, October 22). \emph{Doomscrolling {COVID} news takes an emotional toll -- here's how to make your social media a happier place}. {The Conversation}. \url{http://theconversation.com/doomscrolling-covid-news-takes-an-emotional-toll-heres-how-to-make-your-social-media-a-happier-place-170342}

>>>>

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus?
In what region is it spreading?
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, with use being at an all time high (Statista, 2021).
Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed ``doomscrolling'' emerged (Sharma et al., 2022).
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children.
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health (Sandstrom et al., 2021).
Although research has begun addressing this question
(e.g., Bendau et al., 2021; Eden et al., 2020; Sewall et al., 2021),
it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
This study hence aims to 1) reveal the effect of the different types and channels of social media use on individual well-being, 2) provide generalizable and robust results by analyzing a large-scale longitudinal data-set with 34 waves, and 3) determine the within-person causal effects by analyzing how changes in social media use lead to changes in well-being.

@@

How easily can well-being be affected by external influences?
In general, according to the set-point theory, well-being is surprisingly stable (Lykken, 1999).
Although specific events such as marriage or salary increase can have significant impacts on well-being, in most cases effects are short-term, with well-being after some time returning to prior levels (Diener et al., 2018).
Specific factors such as unemployment, disability, or death, however, can cause long-term changes in well-being (Lucas, 2007).
So although well-being can be affected by external events and factors, this does not happen easily.

@@

Why are the effects of social media use on well-being small on average?
Two prominent media effect theories argue implicitly against strong average negative effects.
First, according to mood management theory (Zillmann, 1988), using media can affect people's moods.
Use can be stimulating or overwhelming, relaxing or boring.
After some time, users implicitly learn which media help them balance their mood and affect according to their own situational needs (Zillmann, 1988).
Those media that eventually become part of one's media repertoire hence, on average, tend to be beneficial for users to regulate their mood (Marciano et al., 2022).
In conclusion, if a certain medium is used frequently, mood-management theory argues that it is likely not detrimental for well-being.

@@

If we look at COVID-19 related use more specifically, how could the various types and channels of COVID-19 related social media use affect well-being?
Several uses and gratifications exist, which help explain why people used social media frequently during the pandemic.
Despite incorrect information, social media provide a vast platform for disseminating \emph{accurate and timely information} about COVID-19 (John Hopkins University, 2023).
Access to reliable information can help people make informed decisions, alleviate uncertainties, and feel empowered during the pandemic.
Social media platforms enable individuals to \emph{connect with others} who are experiencing similar challenges during the pandemic (Guazzini et al., 2022).
Engaging in online communities and support groups can provide emotional support and create a network of like-minded individuals.
Many mental health organizations and professionals utilize social media to share tips, strategies, and resources for \emph{maintaining mental well-being} during the pandemic (Twitter, 2020).
Engaging with such content might help individuals prioritize their mental health and develop resilience during challenging times.
Social media campaigns and initiatives can promote \emph{positive COVID-19 behaviors}, such as mask-wearing, physical distancing, hand hygiene, and vaccination (Athey et al., 2023; Hunt et al., 2022).
Public health organizations and influencers leverage the power of social media to spread awareness and encourage responsible actions, contributing to public health efforts and fostering a sense of collective responsibility.

On the other hand, the effects might be negative, perhaps best explained by the following five mechanisms.
Social media platforms can easily spread \emph{false or misleading information} about COVID-19 (Li et al., 2020).
Due to the ease of sharing and the lack of fact-checking, inaccurate information can go viral and might cause confusion, anxiety, and panic among users.
Constant exposure to COVID-19-related content on social media can lead to \emph{information overload} and contribute to heightened anxiety levels (J. Fan \& Smith, 2021).
The rapid spread of news, updates, and opinions can be overwhelming and might exacerbate existing stress or fears about the pandemic (Sharma et al., 2022).
--\textgreater{}
Social media platforms are known for fostering negativity, with users sometimes engaging in \emph{cyberbullying and harassment}.
Discussions around COVID-19 can become heated and polarized, leading to personal attacks and online conflicts.
Such experiences threaten mental well-being and might contribute to feelings of distress and isolation.
Social media often showcase the highlights and accomplishments of others, encouraging \emph{social comparison} (Przybylski et al., 2013).
During a pandemic, seeing posts about others' successes or seemingly perfect lives might intensify feelings of inadequacy or FOMO, especially when individuals are unable to participate in similar activities due to restrictions or personal circumstances (Sharma et al., 2022).

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Echoing the theoretical rationales outlines above, studies have yielded mixed results.
Some studies found negative effects, indicating that excessive social media use for COVID-19 news led to compulsive behavior and increased stress levels, particularly due to upward social comparison (Stainback et al., 2020; Yue et al., 2022).
Individuals who relied on social media as their primary information source reported higher levels of anxiety and depression symptoms (Bendau et al., 2021).
Similarly, increased COVID-19-related media consumption was associated with higher psychological distress.
On the other hand, some studies reported positive outcomes.
Certain individuals experienced increased virtual community and social connectedness during the pandemic through social media, which contributed to their well-being (Guazzini et al., 2022).
Additionally, engaging more on social media was associated with reduced feelings of loneliness (Latikka et al., 2022).
Several studies reported mostly neutral effects of social media use on well-being indicators (Eden et al., 2020; Sewall et al., 2021).
Overall, the literature demonstrates a mixed picture, highlighting both positive and negative effects of social media use focused on or during COVID-19 on well-being.

@@

When using longitudinal designs to analyze causality, it is important to (a) focus on within-person effects (Hamaker, 2014); to (b) control for confounders (Rohrer \& Murayama, 2023); and to (c) test a plausible interval between measures (Dormann \& Griffin, 2015).
First, in non-experimental designs it makes much sense to analyze causal effects from an internal, within-person perspective (Hamaker, 2014).
If a specific person changes their media diet, we need to measure how this behavior affects their well-being.
Between-person comparisons from longitudinal data cannot provide such insights (Hamaker, 2014).
To test the hypothesis, I thus consider only the within-person effects.

Second, to identify confounders we should control for variables that affect both media use and well-being, which helps isolate the actual effect (Rohrer, 2018).
Because we are adopting a within-person perspective, we need to implement \emph{time-varying} confounders (Rohrer \& Murayama, 2023).
And because we are determining the \emph{overall} causal effect, we need to make sure \emph{not} to control for mediating variables (Rohrer, 2018), for doing so would bias our assessment of the causal effect.
In this study, I hence preregistered to control for the following variables, which either have already been shown or are likely to affect both social media use and well-being, and which also are not mediators:
gender, age, education, Austria country of birth, Austria country of birth of parents, residency Vienna, text-based news consumption, video-based news consumption, household size, health, living space, access to garden, access to balcony, employment, work hours per week, being in home-office, household income, outdoor activities, disposition to take risks, and locus of control (Eger \& Maridal, 2015).

@@

Life satisfaction was measured with the item ``All things considered, how satisfied are you with your life as a whole nowadays?'', which comes from the European Social Survey.
The response options ranged from 0 (\emph{extremely dissatisfied}) to 10 (\emph{extremely satisfied}).

@@

These specific observations notwithstanding, several general trends can be observed.
First, overall the results do suggest that effects of COVID-19 related social media use on well-being tend to take place in the negative as opposed to the positive spectrum.
Although very small, five statistically significant negative results of COVID-19 related social media use on well-being were found.
Only one positive effect emerged.

Second, six significant outcomes emerged for positive or negative affect, but only one for life satisfaction.
Life satisfaction is more stable and not that easily affected by any type or channel of social media communication.
The more fluctuating positive and negative affect, however, were affected (albeit only slightly).
Liking, sharing, and posting COVID-19 related content, and spending more time on Twitter and YouTube to browse COVID-19 related content, all slightly negatively influenced affect.
This is aligned with prior findings which showed that social media use is associated with increased negative affect but not with life satisfaction (Meier \& Reinecke, 2020).
Conversations about COVID-19 on social media are often extreme, negative, or aggressive (L. Fan et al., 2020).
More deeply engaging with this type of content could negatively affect active authors.
The hypothesis that tonality could explain the negative effects is especially supported by the observation that spending more time on Twitter and YouTube than usual increased negative affect.
Communication on both channels is often found to be negative and impolite (e.g., Mueller \& Saeltzer, 2022).
Consuming more negative and misleading information could hence explain the (slightly) increased levels of negative affect.

Third, the results show that it makes sense to analyze different communication types and communication channels.
Reading slightly reduced positive affect, while liking, sharing, and posting slightly increased negative affect.
Interestingly, posting COVID-19 related comment slightly increased negative affect, while at the same time it also slightly increased positive affect.
Posting content is often met with strong reaction, both positive by means of likes and negative by means of critical comments.
Overall, though, posting led to slightly reduced levels of life satisfaction.
In conclusion, whereas it was often stated that passive use is bad and active use good (Verduyn et al., 2015), this pattern was only partially found here.
The results are aligned with the findings from Valkenburg et al. (2022), who could not confirm that active use is good and that passive use is bad.
Focusing on communication channels, Twitter and YouTube seem to be more negative, while Instagram, WhatsApp, and Facebook were neutral.
But, again, all of these effects are very small.

Taken together, the results are hence aligned with the underlying theoretical models and prior empirical results.
The findings support the differential susceptibility of media effects model (Valkenburg \& Peter, 2013), such that effects are generally small and that they depend on the type and channel of communication.
Additional analyses did not reveal that effects depended on gender.
Age also large did not play a significant moderation role, but effects of posting COVID-19 related content were found to be more negative for Generation Z.
Indeed, it has often been argued that effects of social media use are more negative for Gen Z than for prior generations, and this finding can be seen a further tentative support for this hypothesis.
From a broader perspective, the results are well-aligned with mood management theory (Zillmann, 1988) and the uses and gratifications approach (Katz et al., 1973), whose premises preclude particularly negative effects of routine and widespread media consumption.
Both theories posit that if the effects of social media were indeed profoundly negative on average, then people likely would not spend so much time on social media engaging with COVID-19 content.
Finally, recent empirical studies and meta-analyses reported rather small negative effects, too (Meier \& Reinecke, 2020),
echoing the results obtained here.

@@

Focusing on within-person effects and controlling for several potential confounders, this study provides an improved perspective on assessing causality.
However, several challenges remain.
In order to correctly establish causality in non-experimental designs, it is necessary to control for \emph{all} relevant confounding third variables (Rohrer, 2018).
Although this study included are large list of confounders, it could still be that crucial variables were missed.
More thought needs to be invested in which factors to control for and, equally important, for which factors not to control for.

@@

Both media use and well-being were measured using self-reports.
Because assessing well-being necessarily requires introspection, using self-reports for affect and life satisfaction is adequate.
However, for social media use objective measures are preferable, as people often cannot reliably estimate their use (Scharkow, 2016).

Being collected in a single country, the generalizability of the results is limited.
They might not hold true in other cultures, especially non-Western cultures with a different media landscape or alternative social media channels.

@@

In this study, COVID-19 related social media use did not meaningfully affect well-being.
Very small negative effects were found for writing COVID-19 related posts, sharing COVID-19 related content, and spending more time than usual on Twitter.
Factors other than social media use, however, were meaningfully related to well-being, including physical health, exercise, satisfaction with democracy, or believing that one is in control of one's life.
In light of the overall very small effects, engaging in COVID 19-related social media use should not be considered a major concern for one's well-being.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-fanInformationOverloadWellbeing2021}{}}%
Fan, J., \& Smith, A. P. (2021). Information overload, wellbeing and {COVID-19}: {A} survey in {China}. \emph{Behavioral Sciences}, \emph{11}(5), 62. \url{https://doi.org/10.3390/bs11050062}

\leavevmode\vadjust pre{\hypertarget{ref-fanStigmatizationSocialMedia2020}{}}%
Fan, L., Yu, H., \& Yin, Z. (2020). Stigmatization in social media: {Documenting} and analyzing hate speech for {\textsc{COVID}} -19 on {Twitter}. \emph{Proceedings of the Association for Information Science and Technology}, \emph{57}(1). \url{https://doi.org/10.1002/pra2.313}

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-huntSocialMediaBased2022}{}}%
Hunt, I. de V., Dunn, T., Mahoney, M., Chen, M., Nava, V., \& Linos, E. (2022). A social media-based public health campaign encouraging {COVID-19} vaccination across the {United States}. \emph{American Journal of Public Health}, \emph{112}(9), 1253--1256. \url{https://doi.org/10.2105/AJPH.2022.306934}

\leavevmode\vadjust pre{\hypertarget{ref-johnhopkinsuniversityCOVID19Map2023}{}}%
John Hopkins University. (2023). {COVID-19 Map}. In \emph{Johns Hopkins Coronavirus Resource Center}. https://coronavirus.jhu.edu/map.html.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kittelAustrianCoronaPanel2020}{}}%
Kittel, B., Kritzinger, S., Boomgaarden, H., Prainsack, B., Eberl, J.-M., Kalleitner, F., Lebernegg, N. S., Partheymüller, J., Plescia, C., Schiestl, D. W., \& Schlogl, L. (2020). \emph{Austrian {Corona Panel Project} ({SUF} edition)}. {AUSSDA}. \url{https://doi.org/10.11587/28KQNS}

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-normanInterpretationChangesHealthrelated2003}{}}%
Norman, G., Sloan, J., \& Wyrwich, K. (2003). Interpretation of changes in health-related quality of life: {The} remarkable universality of half a standard deviation. \emph{Medical Care}, \emph{41}(5), 582--592.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rohrerTheseAreNot2023}{}}%
Rohrer, J. M., \& Murayama, K. (2023). These are not the effects you are looking for: {Causality} and the within-/between-persons distinction in longitudinal data analysis. \emph{Advances in Methods and Practices in Psychological Science}, \emph{6}(1), 25152459221140842. \url{https://doi.org/10.1177/25152459221140842}

\leavevmode\vadjust pre{\hypertarget{ref-sandstromDoomscrollingCOVIDNews2021}{}}%
Sandstrom, G., Buchanan, K., Aknin, L., \& Lotun, S. (2021). Doomscrolling {COVID} news takes an emotional toll \textendash{} here's how to make your social media a happier place. In \emph{The Conversation}. http://theconversation.com/doomscrolling-covid-news-takes-an-emotional-toll-heres-how-to-make-your-social-media-a-happier-place-170342.

\leavevmode\vadjust pre{\hypertarget{ref-scharkowAccuracySelfreportedInternet2016}{}}%
Scharkow, M. (2016). The accuracy of self-reported {Internet} use\textemdash{{A}} validation study using client log data. \emph{Communication Methods and Measures}, \emph{10}(1), 13--27. \url{https://doi.org/10.1080/19312458.2015.1118446}
